---
title: "Editing"
description: >
  Editing taxonomic databases in Darwin Core format
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Editing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

dwctaxon has two major purposes, (1) editing and (2) validation of taxonomic data in [Darwin Core (DWC)](https://dwc.tdwg.org/terms/#taxon) format. This vignette is about the first. Although you could use dwctaxon to build a taxonomic database from scratch, it is more likely you will be using it to modify an existing database, so we will focus on that kind of use-case.

We start by loading packages needed for this vignette:

```{r setup}
library(dwctaxon)
library(tibble) # recommended for pretty-printing of tibbles
```

## The data

dwctaxon comes with an example dataset, taxonomic data of filmy ferns (family Hymenophyllaceae). Let's take a quick look at the data:

```{r filmy-data}
dct_filmies
```

Although DWC taxon format includes a large number of terms (columns)^[See `dct_terms` for a list], a typical database does not use all of them. `dct_filmies` only includes `r ncol(dct_filmies)` columns. Their usage should mostly be clear to someone familiar with taxonomy, but two columns need more explanation. `taxonID` is a unique ID for each row (name), and `acceptedNameUsageID` is only provided for synonyms; it indicates the `taxonID` of the accepted name. For more information on DWC taxon format, see `vignette("what-is-dwc")`.

The rest of the vignette will consist of modifying this dataset.

## Adding rows

(coming soon!)

## Modifying rows

(coming soon!)

## Filling columns

(coming soon!)
