---
title: "Real World Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real World Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Increase width for printing tibbles
old <- options(width = 140)
```

This vignette demonstrates using dwctaxon on "real life" data found in the wild. Our goal is to import the data and validate it.

First, load the packages used for this vignette.

```{r setup, message = FALSE}
library(dwctaxon)
library(readr)
library(tibble)
library(dplyr)
```

## Import data

We will use the [Database of Vascular Plants of Canada (VASCAN)](http://data.canadensys.net/ipt/resource.do?r=vascan), which is available as a Darwin Core Archive.

The data can be obtained manually by going to the [VASCAN website](http://data.canadensys.net/ipt/resource.do?r=vascan), downloading the Darwin Core Archive, and unzipping it^[If you download the data manually, it may be a different version than the one used here, v37.12].

Alternatively, it can be downloaded and unzipped with R:

```{r download-unzip}
# Set up folders:
# - Specify temporary folder for downloading data
temp_dir <- tempdir()
# - Set name of zip file
temp_zip <- paste0(temp_dir, "/dwca-vascan.zip")
# - Set name of unzipped folder
temp_unzip <- paste0(temp_dir, "/dwca-vascan")

# Download data
download.file(
  url = "https://data.canadensys.net/ipt/archive.do?r=vascan&v=37.12",
  destfile = temp_zip, mode = "wb"
)

# Unzip
unzip(temp_zip, exdir = temp_unzip)

# Check the contents of the unzipped data (the Darwin Core Archive)
list.files(temp_unzip)
```

Next, load the taxonomic data (`taxon.txt`) into R. It is a tab-separated text file, so we use `readr::read_tsv()` to load it.

```{r load-data}
vascan <- read_tsv(paste0(temp_unzip, "/taxon.txt"))

# OK to delete the tempo

# Take a peak at the data
vascan
```

The dataset includes `r nrow(vascan)` rows (taxa) and `r ncol(vascan)` columns.

# Validation

Let's see if the dataset passes validation with dwctaxon.

It is usually a good idea to just run `dct_validate()` with default settings the first time.
If it passes, you can move on.

```{r validation, error = TRUE}
dct_validate(vascan)
```

Looks like we've got problems...

To dig into these in more detail, let's run `dct_validate()` again, but this time output a summary of errors.

```{r validation-summary}
validation_res <- dct_validate(vascan, on_fail = "summary")

validation_res
```

The summary lists one `taxonID` per row. Let's count these to get a higher-level view of what's going on.

```{r summary-analysis}
validation_res %>%
  count(check, error)
```

```{r summary-analysis-hide, show = FALSE, echo = FALSE}
validation_res_sum <-
  validation_res %>%
  count(check, error)

n_error_types <- nrow(validation_res_sum) %>%
  english::english()

n_bad_cols <- validation_res_sum %>%
  filter(error == "Invalid column names detected: id") %>%
  pull(n) %>%
  english::english()

n_bad_sci_name <- validation_res_sum %>%
  filter(error == "scientificName detected with duplicated value") %>%
  pull(n)
```

We see there are `r n_error_types` kinds of errors. 
There is `r n_bad_cols` column with an invalid name (`id`) and `r n_bad_sci_name` rows with duplicated scientific names.

## Investigate errors

### Duplicate names

Let's take a closer look at some of those duplicated names.

```{r check-sci-name-dups}
dup_names <-
  validation_res %>%
  filter(grepl("scientificName detected with duplicated value", error)) %>%
  arrange(scientificName)

dup_names
```

We can join back to the original data to investigate these names.

```{r check-sci-name-dups-orig}
inner_join(
  select(dup_names, taxonID),
  vascan,
  by = "taxonID"
) %>%
  # Just look at the first 6 columns
  select(1:6)
```

We see that in some cases, multiple entries for the exact same scientific name (for example, `Arnica monocephala Rydberg`) differ only in the value for `nameAccordingToID`.

So this seems like something the database manager should fix.

### Invalid column names

Let's see what is in the `id` column.

```{r inspect-id}
vascan %>%
  select(id)

n_distinct(vascan$id)
```

`id` contains numbers that are all unique.
In other words, these appear to be unique key values to each row in the dataset (as one would expect from the name `id`).

It is probably the case that this dataset has a good reason for using the `id` column, even though it is not a standard DwC column.

## Fixing the data

Let's see if we can get this dataset to pass validation.

First, let's remove the duplicated names.
This is something that should be done with more thought, but here let's just keep the first name of each pair.

```{r fix-data}
vascan_fixed <-
  vascan %>%
  filter(!duplicated(scientificName))
```

Next, we will run validation again, but this time allow `id` as an extra column.

```{r validation-2}
dct_validate(
  vascan_fixed,
  extra_cols = "id"
)
```

It passes, so we have now confirmed that the only steps needed to obtain correctly formatted DwC data are to de-duplicate the species names and account for the `id` column.

## Summary

This vignette shows how dwctaxon can be used on DwC data to find possible problems in a taxonomic dataset.
We were able to identify several rows with duplicated scientific names and one column that does not follow DwC standards.
Other than that, it passes validation, giving us confidence that this dataset can be used for downstream taxonomic analyses.

```{r, include = FALSE}
# Reset options
options(old)
```
